Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-26 paper/manuscript.tex
make track changes version
Update manuscript text around updated results

It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high,haddox2018mapping}.
However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
Here, we measure the effects on viral growth in cell culture of all mutations to the HA of a recent human H3N2 strain.
We show that these experimental measurements can help discriminate evolutionarily successful mutations from those found in strains that quickly die out.
However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
The empirical $P$-value represents the result of 1,000 permutations.
}
\end{SCfigure*}

@@

\subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution, and only a weak correlation with mutations from unpassaged viral isolates.
Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.

@@

\matmethods{
\subsection*{Data and computer code}
Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
Computer code used to analyze the data and produce the results in the paper are on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.

@@

To titer the GFP-carrying viruses, we plated $1 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells per well in 12-well plates in IGM.
Four hours after plating, we infected cells with dilutions of viral supernatant.
At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.

}

\showmatmethods{} % Display the Materials and Methods section

>>>>

It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high,haddox2018mapping}.
However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
Here, we measure the effects on viral growth in cell culture of all mutations to the HA of a recent human H3N2 strain.
We show that these experimental measurements can help discriminate evolutionarily successful mutations from those found in strains that quickly die out.
However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are less informative for understanding the evolutionary fate of H3 viral strains.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are less informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
The empirical $P$-value represents the result of 1,000 permutations.
}
\end{SCfigure*}

@@

\subsection*{Measurements made on an H1 HA are less informative for understanding the evolution of H3 influenza}
To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (Figure~\ref{fig:muteffect_maxfreq}).
Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.

@@

{\small
\section*{Materials and Methods}
\subsection*{Data and computer code}
Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
Computer code used to analyze the data and produce the results in the paper are on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.

@@

To titer the GFP-carrying viruses, we plated $1 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells per well in 12-well plates in IGM.
Four hours after plating, we infected cells with dilutions of viral supernatant.
At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.
}
